---
title: "HW2 Econometrics 3"
author: "Matthew Aaron Looney"
date: "10/04/2017"
output: pdf_document
---

```{r HW2 Problem 1 data import, echo=FALSE, message=FALSE, results='asis', cache=F}

cat("\014")
rm(list=ls())
set.seed(12345)

library(readr)
library(gvlma)
library(lmtest)
library(stargazer)

require(foreign)
require(nnet)
require(ggplot2)
require(reshape2)

coffee <- read_csv("~/Google Drive/digitalLibrary/*Econometrics3/Econometrics3/HW2_R_code/GROCERY.csv", col_names = TRUE)

week1379 <- subset(coffee, WEEK==1379 & city=="Pittsfield" & VOL_EQ != 0)

price <- data.frame(price= week1379$DOLLARS/ (week1379$UNITs* week1379$VOL_EQ))

week1379_Pitts_price <- cbind(week1379, price) 



```


```{r HW2 Problem 2 data import, echo=FALSE, message=FALSE, results='asis', cache=F, warning=FALSE}

cat("\014")
rm(list=ls())
set.seed(12345)

library(readr)
library(gvlma)
library(lmtest)
library(stargazer)
library(texreg)

require(foreign)
require(nnet)
require(ggplot2)
require(reshape2)
library(censReg)
library(sampleSelection)

credit <- read_csv("~/Google Drive/digitalLibrary/*Econometrics3/Econometrics3/HW2_R_code/CreditScoring.csv", col_names = TRUE)

credit <- cbind(credit, data.frame(Ln_income=log(credit$INCOME)) )

model <- formula(LOGSPEND~ Ln_income+ AGE+ ADEPCNT+ OWNRENT)

# linear probability model
ols_credit <- lm(model, data=credit)


tobit_left0 <- censReg(model, left=0, data=credit)

partFX_tobit <- margEff(tobit_left0)

heck_formula <- formula(CARDHLDR~ Ln_income+ AGE+ ADEPCNT+ OWNRENT)

heck2Step <- heckit(heck_formula, model, data=credit)
#heckMLE <- selection(heck_formula, model, data=credit)

beta.income.sel <- heck2Step$coefficients[2]
beta.income.out <- heck2Step$coefficients[7]
beta.IMR <- heck2Step$coefficients[11]
delta <- heck2Step$imrDelta

marginal.effect.income <- mean(beta.income.out - beta.income.sel * beta.IMR * delta)


```

\newpage

# Problem 2 

Problem 2. Censoring/Truncation. Greene (2007) analyzed the de- fault behavior and monthly behavior of a large sample of credit card users (13,444).

## (2.1)

Estimate the following model

\[\begin{gathered}
  \log spend = {\beta _1} + {\beta _2}\ln income + {\beta _3}Age + {\beta _4}Adepcnt + \beta 5ownrent + \varepsilon  \hfill \\
   \hfill \\ 
\end{gathered} \]

```{r echo=FALSE, message=FALSE, results='asis', cache=F, warning=FALSE}

stargazer(ols_credit, tobit_left0, heck2Step, header = F, title="Regression output used to answer Problem 2")

```

### (2.1.a)

Using OLS. What is the effect of 10% increase in income on credit card expenditure?

  * Since we are dealing with log-log we can simply multiply the paramater estimate on income by ten, which gives `r ols_credit[["coefficients"]][["Ln_income"]]*10 `. So a 10% increase in income is estimated to increase credit card spending by `r ols_credit[["coefficients"]][["Ln_income"]]*10 `%.
  
### (2.1.b)

Using Censored regression. What is the effect of 10% increase in income on credit card expenditure?

We will need to employ a Censored (Tobit) Regression and calculate the Partial Effects.

The general formulation for the Tobit Model (Greene 7th. ed., pg 848):

\[\begin{gathered}
  {y_i}^* = {x_i}^\prime \beta  + {\varepsilon _i} \hfill \\
  \begin{array}{*{20}{c}}
  {{y_i} = 0} \\ 
  {{y_i} = {y_i}^*} 
\end{array}\left\{ \begin{gathered}
  {\text{if }}{y_i}^* \leqslant 0 \hfill \\
  {\text{if }}{y_i}^* \geqslant 0 \hfill \\ 
\end{gathered}  \right. \hfill \\ 
\end{gathered} \]

The proper Partial Effects formula:

\[\frac{{\partial E[y|x]}}{{\partial x}} = \beta \Pr ob[a < {y^*} < b]\]

Where I compute the partial effect at each observation and then compute the mean.

The parginal effect of Ln_income on LOGSPEND is `r partFX_tobit[1]`. Therefore, a 10% increase of income is estiimated to increase credit card spending by `r 10*partFX_tobit[1]`.

### (2.1.c)

Using Heckman Two-Step Estimator. What the is effect of 10% increase in income on credit card expenditure?


